Global patterns of forest autotrophic carbon fluxes

### 2020-04-07 manuscript/manuscript draft.Rmd
added citations to manuscript draft

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity, while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate ([Cavaleri et al. 2015](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.12860); [Rogelj et al. 2018](https://www.ipcc.ch/site/assets/uploads/sites/2/2019/05/SR15_Chapter2_Low_Res.pdf)). However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

FACF decrease with latitude, but it remains unclear whether and how the shape of this relationship varies across fluxes. Studies agree that FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019](*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*). In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity ($ANPP$), finding evidence of a weak negative relationship between $ANPP$ and latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in FACF is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011]([Hursh et al. 2017](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.13489)). The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. While these fail to capture some important aspects of climate such as seasonality, they do describe broad trends in temperature and water availability, and therefore capture a substantial portion of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. {**It would be good to add some more citations on soil respiraiton. I'm sure BBL can help.**}

@@

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/); Potapov et al. 2008. Mapping the World’s Intact Forest Landscapes by Remote Sensing. Ecology and Society 13 (2), 51. ), so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (** Nobby AFM paper: Understanding the controls over forest carbon use efficiency on small spatial scales: Effects of forest disturbance and tree diversity**) [*purpose for respiration/ other compenents (**See Nobby's comment in manuscript-draft_NK.pdf**)].

@@

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ (*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*) and $ANPP_{woody stem}$ (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*; *[Hubau et al. 2020](https://doi.org/10.1038/s41586-020-2035-0)*), along with some C cycle components not considered here--tree mortality (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*; *[McDowell et al. 2018](https://nph.onlinelibrary.wiley.com/doi/full/10.1111/nph.15027 )*), soil respiration ([Bond Lamberty & Thompson 2010](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature08930/)), and heterotrophic soil respiration ([Bond-Lamberty et al. 2018](https://www-nature-com.smithsonian.idm.oclc.org/articles/s41586-018-0358-x)). However, increasing C flux rates are not universal (e.g., *[Rutishauser et al., 2019](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14833?casa_token=imtB9LD9mL8AAAAA%3AYWgsqNUfRkgs6_dIuDzmwUzL0qVyyxojhbj3jjfRcykBy8ay9T4qMOKPeOPytSGET9muIs8myhQOh3E)*, **MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of cliamte, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

- consistent with [Hofhansl et al. 2015](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1002/2014GB004934) (**verify**) , we found a slight tendency for warmer sites to have higher aboveground allocation

>>>>

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity, while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate [@cavaleri_urgent_2015; @rogelj_mitigation_2018]. However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

FACF decrease with latitude, but it remains unclear whether and how the shape of this relationship varies across fluxes. Studies agree that FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity ($ANPP$), finding evidence of a weak negative relationship between $ANPP$ and latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in FACF is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. While these fail to capture some important aspects of climate such as seasonality, they do describe broad trends in temperature and water availability, and therefore capture a substantial portion of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. {**It would be good to add some more citations on soil respiraiton. I'm sure BBL can help.**}

@@

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration [@kunert_understanding_2019].
[*purpose for respiration/ other compenents* (**See Nobby's comment in manuscript-draft_NK.pdf**)].

@@

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ [@li_mapping_2019] and $ANPP_{woody stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates are not universal [@rutishauser_testing_2020] (**MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of cliamte, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

- consistent with [@hofhansl_new_2015] (**verify**) , we found a slight tendency for warmer sites to have higher aboveground allocation
